Relational Parametricity for Algebraically Indexed Types

### 2012-05-22 paper/geometry-twocolumn.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\documentclass{article}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Geometric Invariance and Relational Parametricity,
            pdfauthor={Robert Atkey},
            pdfkeywords={}}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\usepackage{amssymb}
\usepackage{stmaryrd}

\newcommand{\sepbar}{\mathrel|}

\newcommand{\GA}{\mathrm{GA}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Is}{\mathrm{Is}}
\newcommand{\Orth}{\mathrm{O}}
\newcommand{\SOrth}{\mathrm{SO}}
\newcommand{\Transl}{\mathrm{T}}
\newcommand{\Scale}{\mathrm{Scale}}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}

\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}

\newenvironment{example}{\vspace{0.1in}\noindent\textbf{Example}\quad}{}

\newcommand{\ignore}[1]{}

\title{Geometric Invariance and Relational Parametricity}
\author{Robert Atkey}

\begin{document}

\maketitle

\section{Coordinate-free Geometry and Invariance}

@@

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points in the same frame, but that the
function itself should be invariant with respect to the choice of
frame. We use Reynolds' idea of preservation of relations to formalise
this: for any $F \in \GA(2)$ (i.e.~any choice of frame), the function
$f$ should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\ (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
\end{equation}
Unfolding the definition of $R_{\texttt{pt}}$, we can see that this
statement is equivalent to the following statement expressing
invariance of $f$ with respect to changes of frame: for all $F \in
\GA(2)$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(F\vec{v_1},F\vec{v_2}) = F(f(\vec{v_1},\vec{v_2})).
\end{equation}
So Reynolds' idea of preservation of relations, when instantiated with
the family of relations $R_{\texttt{pt}}$, yields exactly the
geometric property of invariance under change of frame.

@@

\section{A Type System for Geometric Invariance}
\label{sec:type-system-geom-intro}

@@

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation.
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = ((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. Note that the intended
denotation in the second line does not take any argument corresponding
to the universally quantified $F$ in the type. In our calculus,
geometric group expressions are erased at runtime.

@@

\begin{example}
  With just the single primitive operation of affine combination, we
  can already express some useful programs. For example, the
  evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
  two endpoints and a single control point) can be expressed using the
  affine combination primitive and three steps of De Casteljau's
  algorithm:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{quadraticB\acute{e}zier} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
      \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
        & t \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$,
  $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for $t \in
  [0,1]$ gives us the point on the curve at ``time'' $t$. Note that
  the type of $\mathrm{quadraticB\acute{e}zier}$ tells us immediately
  that it preserves all invertible affine maps.
  %FIXME: forward ref to ``free theorem''.
\end{example}

@@

We can now give vector addition a more general type, which
demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation is invariant under invertible linear maps
but negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall B \mathord: \GL(2), t \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t\rangle \to \mathsf{vec}\langle B\ltimes(-t)\rangle
\end{displaymath}
With addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
\end{array}
\end{displaymath}

@@

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthognal transformations and
  arbitrary translations.
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the types. First we have computed the two offset vectors $p_2 -
  p_1$ and $p_3 - p_1$. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we demonstrated above. We then compute the cross product
  of the two vectors to determine the area of the parallelogram
  described by the sides of the triangle. This value has type
  $\mathsf{real}\langle \det O \rangle$. We know that the determinant
  of an orthogonal transformation is always either $-1$ or $1$, so we
  have a result which varies according to possible reflections of the
  space. To remove this variation, we apply the $\mathrm{abs}$
  function to leave us with a value of type $\mathsf{real}\langle
  |\det O| \rangle$. This type is equal to the type
  $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.
\end{example}

% \subsection{Transformations Within and Between Frames}
% \label{sec:runtime-rep-frames}

% \begin{enumerate}
% \item A translation with respect to some frame is simply a vector with
%   respect to that frame's basis. We can apply it to points just by 
% \item 
% \end{enumerate}

% Singleton types

% Representing frames, a type representing translations:
% \begin{displaymath}
%   \mathsf{translation}\langle T \rangle
% \end{displaymath}
% And representing bases, as linear maps
% \begin{displaymath}
%   \mathsf{invLinearMap}\langle B \rangle
% \end{displaymath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIXME: write down all the groups, and the functions between them

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\bibliography{paper}

\end{document}

>>>>

\section{Motivating Example: 2D Geometry}

\subsection{Coordinate-free Geometry and Invariance}

@@

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points in the same frame, but that the
function itself should be invariant with respect to the choice of
frame. We use Reynolds' idea of preservation of relations to formalise
this: for any $F \in \GA(2)$ (i.e.~any choice of frame), the function
$f$ should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \begin{array}{l}
    \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\\
    \hspace{2cm}(f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
  \end{array}
\end{equation}
Unfolding the definition of $R_{\texttt{pt}}$, we can see that this
statement is equivalent to the following statement expressing
invariance of $f$ with respect to changes of frame: for all $F \in
\GA(2)$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(F\vec{v_1},F\vec{v_2}) = F(f(\vec{v_1},\vec{v_2})).
\end{equation}
So Reynolds' idea of preservation of relations, when instantiated with
the family of relations $R_{\texttt{pt}}$, yields exactly the
geometric property of invariance under change of frame.

@@

\subsection{A Type System for Geometric Invariance}
\label{sec:type-system-geom-intro}

@@

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation.
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} :\\
    \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
    \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. Note that the intended
denotation in the second line does not take any argument corresponding
to the universally quantified $F$ in the type. In our calculus,
geometric group expressions are erased at runtime.

@@

\begin{example}
  With just the single primitive operation of affine combination, we
  can already express some useful programs. For example, the
  evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
  two endpoints and a single control point) can be expressed using the
  affine combination primitive and three steps of De Casteljau's
  algorithm:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{quadraticB\acute{e}zier} :\\
      \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
      \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
        & t \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
  $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
  that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
  immediately that it preserves all invertible affine maps.
  %FIXME: forward ref to ``free theorem''.
\end{example}

@@

We can now give vector addition a more general type, which
demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  \begin{array}{l}
    (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \\
    \quad\quad\quad\quad\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
\end{array}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation is invariant under invertible linear maps
but negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall B \mathord: \GL(2), t \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t\rangle \to \mathsf{vec}\langle B\ltimes(-t)\rangle
\end{displaymath}
With addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \\
    \hspace{1cm}\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
\end{array}
\end{displaymath}

@@

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthognal transformations and
  arbitrary translations.
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \\
      \hspace{1.2cm}\mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the types. First we have computed the two offset vectors $p_2 -
  p_1$ and $p_3 - p_1$. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we demonstrated above. We then compute the cross product
  of the two vectors to determine the area of the parallelogram
  described by the sides of the triangle. This value has type
  $\mathsf{real}\langle \det O \rangle$. We know that the determinant
  of an orthogonal transformation is always either $-1$ or $1$, so we
  have a result which varies according to possible reflections of the
  space. To remove this variation, we apply the $\mathrm{abs}$
  function to leave us with a value of type $\mathsf{real}\langle
  |\det O| \rangle$. This type is equal to the type
  $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.
\end{example}
